Retrofitting Effect Handlers onto OCaml

### 2020-11-20 retro-concurrency.tex
Udpates
Updates

We now present the implementation details of effect handlers in
Multicore OCaml. We assume an x86\_64 architecture for the remainder of this paper
-- although  Multicore OCaml currently only supports the x86\_64 backend on
Linux, our design is intended to support other CPUs and OSs.

@@

%It is useful to note that stack overflow is a rare operation
%and the CPU branch predictor will correctly predict the branch with high
%accuracy.

@@

Next we look at benchmarks that utilize non-local control flow using effect
handlers. First, we quantify the overhead of individual operations in effect
handling. Consider the following annotated code:
\begin{lstlisting}
effect E : unit
(* a *) match (* b *) perform E (* d *) with
| v -> (* e *) v
| effect E k -> (* c *) continue k ()
\end{lstlisting}
The sequence \texttt{a-b} involves allocating a new fiber and switching to it.
\texttt{b-c} is performing the effect and handling it. \texttt{c-d} is resuming
the continuation. \texttt{d-e} is returning from the fiber with a value and
freeing the fiber. We measured the time taken to execute these sequences using
|perf| support for cycle-accurate tracing on modern Intel processors. We
executed 10 iterations of the code, with 3 warm-up runs. We observed that the
sequences \texttt{a-b}, \texttt{b-c}, \texttt{c-d} and \texttt{d-e} took
\textbf{23ns, 5ns, 11ns and 7ns}, respectively. For reference, the idle
memory load latency for the local NUMA node is 93.2ns as measured using the
Intel MLC tool~\cite{mlc}.

@@

\begin{figure}
	\begin{minipage}{0.49\linewidth}
		\centering
		\includegraphics[width=0.9\linewidth]{http-benchmarks/throughput.pdf}
		\subcaption{Throughput}
		\label{grf:throughput}
	\end{minipage}
	\begin{minipage}{0.49\linewidth}
		\includegraphics[width=0.9\linewidth]{http-benchmarks/latency.pdf}
		\subcaption{Tail latency}
		\label{grf:latency}
	\end{minipage}
	\vspace{-3mm}
	\caption{Web server performance.}
	\vspace{-5mm}
\end{figure}

@@

Multicore OCaml builds on the existing defensive coding practices against
exceptions to clean up resources on non-returning exits. Our continuations are
assumed to be resumed exactly once using |continue| or |discontinue|. As a
result, when a computation performs an effect, we are guaranteed that the
control will return. For the non-returning cases (value and exceptional
return), the code already handles resource cleanup. OCaml does not have a
\texttt{\footnotesize try/finally} construct commonly used for resource cleanup
in many programming languages. Alternative standard libraries such as
Base~\cite{BaseProtect} and Core~\cite{CoreProtect} provide mechanisms
analogous to |unwind-protect|, which are in turn implemented using exceptions.
Thus, the linear use of continuations enabled by the |discontinue| primitive
ensures backwards compatibility of legacy OCaml systems code under non-local
control flow introduced by effect handlers. Leijen~\cite{Leijen18} explicitly
extends effect handlers with |initially| and |finally| clauses in Koka for
resource safety. Sivaramakrishnan et al.~\cite{TFP17} describe the interaction
of effect handlers and asynchronous exceptions. This is orthogonal to the
contributions of this paper. Our focus is the compiler and runtime system
support for implementing effect handlers.

\vspace{-2mm}
\section{Conclusions}
\label{sec:conc}

@@

@@

Languages with effect handlers typically allow continuations to be resumed more
than once to enable use cases such as backtracking search and probabilistic
programming (Uber's Pyro language~\cite{Pyro}). Multicore OCaml has an
experimental feature for explicitly cloning continuations, which makes a copy
of each of its constituent fibers.

>>>>

We now present the implementation details of effect handlers in Multicore
OCaml. We assume an x86\_64 architecture for the remainder of this paper.
Although Multicore OCaml currently only supports the x86\_64 backend on Linux,
our design does not preclude other CPUs and OSs.

@@

@@

Next we look at benchmarks that utilize non-local control flow using effect
handlers. First, we quantify the overhead of individual operations in effect
handling. Consider the following annotated code:
\begin{lstlisting}
effect E : unit
(* a *) match (* b *) perform E (* d *) with
| v -> (* e *) v
| effect E k -> (* c *) continue k ()
\end{lstlisting}
The sequence \texttt{a-b} involves allocating a new fiber and switching to it.
\texttt{b-c} is performing the effect and handling it. \texttt{c-d} is resuming
the continuation. \texttt{d-e} is returning from the fiber with a value and
freeing the fiber. We measured the time taken to execute these sequences using
|perf| support for cycle-accurate tracing on modern Intel processors. We
executed 10 iterations of the code, with 3 warm-up runs. We observed that the
sequences \texttt{a-b}, \texttt{b-c}, \texttt{c-d} and \texttt{d-e} took
\textbf{23 ns, 5 ns, 11 ns and 7 ns}, respectively. For reference, the idle
memory load latency for the local NUMA node is 93.2 ns as measured using the
Intel MLC tool~\cite{mlc}.

@@

\begin{figure}
	\begin{minipage}{0.49\linewidth}
		\centering
		\includegraphics[width=0.9\linewidth]{http-benchmarks/throughput.pdf}
		\vspace{-2mm}
		\subcaption{Throughput}
		\label{grf:throughput}
	\end{minipage}
	\begin{minipage}{0.49\linewidth}
		\includegraphics[width=0.9\linewidth]{http-benchmarks/latency.pdf}
		\vspace{-2mm}
		\subcaption{Tail latency}
		\label{grf:latency}
	\end{minipage}
	\vspace{-3mm}
	\caption{Web server performance.}
	\vspace{-5mm}
\end{figure}

@@

Multicore OCaml builds on the existing defensive coding practices against
exceptions to clean up resources on non-returning exits. Our continuations are
assumed to be resumed exactly once using |continue| or |discontinue|. As a
result, when a computation performs an effect, we are guaranteed that the
control will return. For the non-returning cases (value and exceptional
return), the code already handles resource cleanup. OCaml does not have a
\texttt{\footnotesize try/finally} construct commonly used for resource cleanup
in many programming languages. Alternative standard libraries such as
Base~\cite{BaseProtect} and Core~\cite{CoreProtect} provide mechanisms
analogous to |unwind-protect|, which are in turn implemented using exceptions.
Thus, the linear use of continuations enabled by the |discontinue| primitive
ensures backwards compatibility of legacy OCaml systems code under non-local
control flow introduced by effect handlers. Leijen~\cite{Leijen18} explicitly
extends effect handlers with |initially| and |finally| clauses in Koka for
resource safety. Dolan et al.~\cite{TFP17} describe the interaction
of effect handlers and asynchronous exceptions. This is orthogonal to the
contributions of this paper. Our focus is the compiler and runtime system
support for implementing effect handlers.

\vspace{-3mm}
\section{Conclusions}
\label{sec:conc}

@@

\if{0}
\subsection*{Dynamic Semantics}

\begin{definition}[Initial configuration]
	Given an expression $e$, we define the initial configuration $\textsf{init}(e)$
	to be $\configa{e}{\emptyset}{\cstacka{[]}{\bullet}}$.
\end{definition}

Intuitively, the initial environment is empty and the initial stack is an empty
C stack.

\begin{definition}[Final value configuration]
	Given a configuration $\config$, we say that the configuration is the final
	value configuration (written $\textsf{finalval}(\config)$) if ~$\config =
	\configa{v}{\env}{\cstacka{[]}{\bullet}}$~ for some value $v$ and environment
	$\env$.
\end{definition}

\begin{definition}[Final exception configuration]
	Given a configuration $\config$, we say that the configuration is the final
	exception configuration (written $\textsf{finalexn}(\config)$) if ~$\config
	= \configa{v}{\env}{\cstacka{\ffun{\exnval{l}}::\fl}{\bullet}}$~ for some
	value $v$, label $l$, frame list $\fl$ and environment $\env$.
\end{definition}

Given a configuration $\config = \configa{\term}{\env}{\stack}$, we define
$\textsf{term}(\config)$ to be $\tau$. We define the relation $\multistep$ as
the transitive reflexive closure of the top-level reduction relation $\step$.

\subsubsection*{Example 1}

Let

\begin{smathpar}
	\begin{array}{rcll}
		e & = & (\olam{x}{ & \kw{match} ~\perform{E}{~0}~ \kw{with} \\
			&   &            & \mid \caseexn{\textsf{Unhandled}}{x}{42} \\
			&   &            & \mid \caseval{v}{v}})~0
	\end{array}
\end{smathpar}

then $init(e) \multistep ~\config \wedge final(\config) \wedge term(\config) =
42$.

\begin{proof}
\end{proof}
\fi

@@

Languages with effect handlers typically allow continuations to be resumed more
than once to enable use cases such as backtracking search and probabilistic
programming (for example, see Uber's Pyro language~\cite{Pyro}). Multicore
OCaml has an experimental feature for explicitly cloning continuations, which
makes a copy of each of its constituent fibers.
